---
title: "Acral melanoma MX - consensus clustering analysis, enrichment and visualization of gene signatures across clusters"
author: "Estef Vazquez"
output:
  html_document:
    theme: simplex
    toc: yes
    toc_float:
      collapsed: true
---

Overview
This document provides code for running consensus clustering to identify molecular subtypes (two-level clustering approach, sample classification followed by gene modules), differential expression, enrichment analysis and visualization of gene expression signatures in 44 acral melanoma samples from Mexican patients. The analysis also generates a heatmap showing expression patterns and boxplot comparisons of differentially expressed genes across clusters.


## Load required libraries

```{r, message=FALSE}

library(here)
library(cola)
library(ComplexHeatmap)
library(clusterProfiler)
library(tidyverse)
library(ggpubr)
library(RColorBrewer)
library(colorRamp2)
library(circlize)
library(cowplot)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(grid)
library(gridExtra)
library(reshape2)
```


## Clustering and signature analysis

```{r}
# Load data 
mt <- readRDS("acralMX_expression_data_update44s.rds")
metadata <- read_csv("acralMX_metadata_cc_update.csv")

# Prepare annotation data 
metadata <- as.data.frame(metadata)
cc_anno <- metadata %>%
  select(sample, TMB, WGD) %>%
  mutate(WGD = factor(WGD)) %>%
  as.data.frame()

# Set rownames and remove sample column
rownames(cc_anno) <- cc_anno$sample
cc_anno <- cc_anno[, -1]

# Define colors 
colors <- list('WGD' = c('WGD' = 'deeppink4', 'no_WGD' = 'cadetblue1'),
               'TMB' = colorRamp2(c(0, 600), c("white", "red")))

# Dimensions 
dim(mt)
dim(cc_anno)

# Run consensus clustering with multiple methods
consensus_results <- run_all_consensus_partition_methods(
  mt, 
  anno = cc_anno, 
  anno_col = colors
)

# Generate report 
cola_report(consensus_results, output_dir= "consensus_clustering_report")

# Extract results for the best performing combination (top-value and partitioning)
SD_PAM_K3 = consensus_results["SD", "pam"]

# Analyze stats and visualize 
get_stats(consensus_results, k = 3)

consensus_heatmap(consensus_results["SD:pam"], k = 3)
collect_classes(consensus_results["SD:pam"])

# PCA
dimension_reduction(SD_PAM_K3, k = 3, method = "PCA")

# Membership heatmap
membership_heatmap(SD_PAM_K3, k = 3)

# Test association with WGD
test_to_known_factors(consensus_results)

# Extract sample cluster classification
classification <- cbind(get_classes(consensus_results["SD:pam"], k = 3), 
                        get_membership(consensus_results["SD:pam"], k = 3))
write.table(classification, file = "cc_classification.tsv", sep = "\t",  col.names=NA)

# Fix levels for visualization 
split <- factor(paste0("Cluster\n", classification$class), levels=c("Cluster\n1","Cluster\n2","Cluster\n3"))


# Gene Signature analysis
# Identify differentially expressed genes using F-test
DE_signatures <- get_signatures(SD_PAM_K3, k = 3)
table(DE_signatures$km)


# Create Gene ID mapping 
gene_mapping <- data.frame(
  which_row = 1:nrow(mt),  
  ensembl_id = rownames(mt),  
  stringsAsFactors = FALSE
)

gene_mapping$gene_symbol <- mapIds(
  org.Hs.eg.db,
  keys = gene_mapping$ensembl_id,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

# Join 


DE_genes_with_ids <- DE_signatures %>%
  left_join(gene_mapping, by = "which_row") %>%
  select(which_row, ensembl_id, gene_symbol, fdr, p_value, everything())

# Determine cluster with highest expression
DE_signatures_fc <- de_genes_with_ids %>%
  rowwise() %>%
  mutate(
    highest_expr_cluster = which.max(c(mean_1, mean_2, mean_3)),
    # Calculate  maximum difference between cluster means 
max_mean_diff = max(c(mean_1, mean_2, mean_3)) - min(c(mean_1, mean_2, mean_3))
) %>%
ungroup()

write.csv(DE_signatures_fc, "DE_gene_signatures.csv")


# Functional Enrichment Analysis
enrichment_results = functional_enrichment(object = SD_PAM_K3, k = 3, ontology = c("BP", "MF", "CC"))

# Extract gene lists for each module
genes_group1 <- rownames(mt)[DE_signatures$which_row[DE_signatures$km == 1]]
genes_group2 <- rownames(mt)[DE_signatures$which_row[DE_signatures$km == 2]]
genes_group3 <- rownames(mt)[DE_signatures$which_row[DE_signatures$km == 3]]

# Map ensembl id to gene symbols for each module
map_ids_to_symbols <- function(ensembl_ids) {
  ensembl_ids <- sub("\\.[0-9]+$", "", ensembl_ids)
  gene_symbols <- mapIds(org.Hs.eg.db, 
                        keys = ensembl_ids,
                        column = "SYMBOL", 
                        keytype = "ENSEMBL",
                        multiVals = "first")
  return(gene_symbols)
}
all_genes_group1 <- map_ids_to_symbols(genes_group1)
all_genes_group2 <- map_ids_to_symbols(genes_group2)
all_genes_group3 <- map_ids_to_symbols(genes_group3)


# Run GO enrichment analysis for each module
enrichBP_1 <- enrichGO(
  gene = all_genes_group1, 
  OrgDb = org.Hs.eg.db,
  keyType = "SYMBOL",  
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05
)

enrichBP_2 <- enrichGO(
  gene = all_genes_group2, 
  OrgDb = org.Hs.eg.db,
  keyType = "SYMBOL",
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05)

enrichBP_3 <- enrichGO(
  gene = all_genes_group3, 
  OrgDb = org.Hs.eg.db,
  keyType = "SYMBOL",
  ont = "BP",
  pvalueCutoff = 0.05
)

BP1 <- as.data.frame(enrichBP_1)
BP2 <- as.data.frame(enrichBP_2)
BP3 <- as.data.frame(enrichBP_3)

# Create enrichment plots
dotplot(enrichBP_1, showCategory = 15, title = "Gene Module 1 - Biological Processes")
dotplot(enrichBP_2, showCategory = 15, title = "Gene Module 2 - Biological Processes")
dotplot(enrichBP_3, showCategory = 15, title = "Gene Module 3 - Biological Processes")

# Calculate average scaled means 
group1_means <- colMeans(DE_signatures[DE_signatures$km == 1, c("scaled_mean_1", "scaled_mean_2", "scaled_mean_3")])
group2_means <- colMeans(DE_signatures[DE_signatures$km == 2, c("scaled_mean_1", "scaled_mean_2", "scaled_mean_3")])
group3_means <- colMeans(DE_signatures[DE_signatures$km == 3, c("scaled_mean_1", "scaled_mean_2", "scaled_mean_3")])

# Calculate and visualize expression patterns
group1_means <- colMeans(DE_signatures[DE_signatures$km == 1, c("scaled_mean_1", "scaled_mean_2", "scaled_mean_3")])
group2_means <- colMeans(DE_signatures[DE_signatures$km == 2, c("scaled_mean_1", "scaled_mean_2", "scaled_mean_3")])
group3_means <- colMeans(DE_signatures[DE_signatures$km == 3, c("scaled_mean_1", "scaled_mean_2", "scaled_mean_3")])

pattern_df <- data.frame(
  Cluster = rep(c("1", "2", "3"), 3),
  Expression = c(group1_means, group2_means, group3_means),
  SignatureGroup = rep(c("Gene Module 1", "Gene Module 2", "Gene Module 3"), each = 3)
)
expression_pattern_plot <- ggplot(pattern_df, aes(x = Cluster, y = Expression, group = SignatureGroup, color = SignatureGroup)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 3) +
  theme_classic() +
  labs(
    title = "Gene Expression Patterns Across Sample Clusters",
    x = "Sample Cluster",
    y = "Scaled Mean Expression"
  ) +
  theme(legend.position = "bottom")

ggsave("gene_expression_patterns.pdf", expression_pattern_plot, width = 10, height = 6)


#Prepare Data for Visualization 
# Extract signature gene expression data
mt_genes <- rownames(SD_PAM_K3)
sig_index <- DE_signatures$which_row

# Create gene signature matrix
df <- data.frame(chr=SD_PAM_K3@top_value_list, stringsAsFactors=FALSE)
gene_list <- rownames_to_column(df, var = "ensembl")
gene_signature <- gene_list[sig_index,]
gene_signature <- gene_signature[,-2]
subset_signature <- mt[rownames(mt) %in% gene_signature, ]  
dim(subset_signature)

# Z-score normalization
data_subset_signatures_z <- t(scale(t(subset_signature)))

#  Or Load the precomputed z-score matrix
data_subset_signatures_z <- readRDS("DE_signatures_zscores.rds")

if (file.exists("DE_signatures_zscores.rds")) {
  data_subset_signatures_z <- readRDS("DE_signatures_zscores.rds")
} else {
  data_subset_signatures_z <- t(scale(t(subset_signature)))
  saveRDS(data_subset_signatures_z, "DE_signatures_zscores.rds")
}

```


## Heatmap visualization

```{r}

# Verify samples in mt and metadata are present
all_expr_samples_in_metadata <- all(colnames(data_subset_signatures_z) %in% rownames(metadata))
print(paste("All expression samples in metadata:", all_expr_samples_in_metadata))
all_metadata_samples_in_expr <- all(rownames(metadata) %in% colnames(data_subset_signatures_z))
print(paste("All metadata samples in expression data:", all_metadata_samples_in_expr))

# Verify order 
metadata_ordered <- metadata[colnames(data_subset_signatures_z), ]
identical(rownames(metadata_ordered), colnames(data_subset_signatures_z))

# Samples 
samples_ht <- c("PR51978a", "PR51948a", "PR40984a", "PR40966a", "PR41905a", "PR41916a", "PR51951a", "PR40963a", "PR41045a", "PR40996a", "PR51979a", 
                 "PR41023f", "PR40986a", "PR41912a", "PR41928d", "PR51928a", "PR41017a", "PR41901a", "PR40985a", "PR41044a", "PR51932a", "PR41011a",
                 "PR41930d", "PR40989a", "PR41927d", "PR40971a", "PR40962a", "PR40961a", "PR41920a", "PR40967a", "PR41030a", "PR40994a", "PR41020a",     
                 "PR40997a", "PR41921a", "PR40968a", "PR41002a", "PR41913a", "PR40969a", "PR41025a", "PR40988a", "PR41035a", "PR40970a", "PR40990a")

# Verify class order
classification_ordered <- classification[samples_ht, ]
table(classification_ordered$class)
nrow(classification_ordered)
length(samples_ht) 

# Create split
split <- factor(paste0("Cluster", classification_ordered$class), 
                levels=c("Cluster1", "Cluster2", "Cluster3"))
names(split) <- samples_ht


# Verify data is aligned
data_subset_signatures_z <- data_subset_signatures_z[, samples_ht]
identical(names(split), samples_ht)  
identical(names(split), colnames(data_subset_signatures_z))  
metadata_ordered <- metadata_ordered[samples_ht, ]

identical(rownames(metadata_ordered), samples_ht)  
identical(rownames(metadata_ordered), colnames(data_subset_signatures_z)) 


# Annotations 
# TMB
tmb_breaks <- c(min(metadata_ordered$TMB, na.rm=TRUE), 
                median(metadata_ordered$TMB, na.rm=TRUE), 
                max(metadata_ordered$TMB, na.rm=TRUE))

col_fun_tmb = colorRamp2(tmb_breaks, c("white", "#fdf542", "#ff0000"))


# Age
age_breaks <- c(min(metadata_ordered$age, na.rm=TRUE),
                median(metadata_ordered$age, na.rm=TRUE),
                max(metadata_ordered$age, na.rm=TRUE))

col_fun_age = colorRamp2(age_breaks, c("#FFEDAB", "#FFA500", "#CD6600"))


# Tumor purity
col_fun_purity = colorRamp2(c(0, 0.5, 1.0), c("white", "#fdf542", "#ff0000"))


# Colors HT
mycol <- colorRampPalette(c('green', 'black', 'purple'))(100)
mybreaks <- seq(-2, 2, length.out = 100) 

colors <- list(
  'mutation_status' = c('BRAF' = '#F48FB1', 'NF1' = '#81C784', 
                        'NRAS' = '#FFAB91', 'KIT' = '#80DEEA',
                        'Other' = '#BDBDBD', 'multihit' = '#FFF176'), 
  'stage' = c('0' = '#E0E0E0', '1' = '#BBDEFB', '2' = '#90CAF9',
              '3' = '#5C6BC0', '4' = '#4527A0'),
  'sex' = c('F' = '#4DAF4A', 'M' = '#FC4E2A'),
  'age' = col_fun_age,
  'ulceration' = c('1' = '#CB181D', '0' = '#2196F3'),
  'WGD' = c('1' = 'deeppink4', '0' = '#4FC3F7'),
  'TMB' = col_fun_tmb,
  'tumor_purity' = col_fun_purity
)


colAnn <- HeatmapAnnotation(
  df = metadata_ordered,
  which = 'col', 
  na_col = "black", 
  col = colors,
  simple_anno_size = unit(0.5, "cm"),
  annotation_width = unit(c(1, 8), 'cm'),  # Adjusted for 8 columns
  gap = unit(0.5, 'mm'),
  annotation_height = unit(0.5, 'cm'),  # Slightly reduced for fewer columns
  annotation_name_gp = gpar(fontsize = 15),
  annotation_name_align = TRUE,
  annotation_name_side = 'left',
  annotation_legend_param = list(
      mutation_status = list(
        title = 'Mutation Status',
        title_gp = gpar(fontsize = 22, fontface = 'bold'),
        labels_gp = gpar(fontsize = 20)),
      stage = list(
        title = 'Stage',
        title_gp = gpar(fontsize = 22, fontface = 'bold'),
        labels_gp = gpar(fontsize = 20)),
      sex = list(
        title = 'Sex',
        title_gp = gpar(fontsize = 22, fontface = 'bold'),
        labels_gp = gpar(fontsize = 20)),
      age = list(
        title = 'Age',
        title_gp = gpar(fontsize = 22, fontface = 'bold'),
        labels_gp = gpar(fontsize = 20)),
      ulceration = list(
        title = 'Ulceration',
        at = c("1", "0"),  # Adapt based on your actual values
        labels = c("Yes", "No"),
        title_gp = gpar(fontsize = 22, fontface = 'bold'),
        labels_gp = gpar(fontsize = 20)),
      WGD = list(
        title = 'WGD',
        at = c("1", "0"),  # Adapt based on your actual values
        labels = c("Yes", "No"),
        title_gp = gpar(fontsize = 22, fontface = 'bold'),
        labels_gp = gpar(fontsize = 20)),
      TMB = list(
        title = 'TMB',
        title_gp = gpar(fontsize = 22, fontface = 'bold'),
        labels_gp = gpar(fontsize = 20)),
      tumor_purity = list(
        title = 'Tumor Purity',
        title_gp = gpar(fontsize = 22, fontface = 'bold'),
        labels_gp = gpar(fontsize = 20))
  )
)
```


# Clustering DE signatures - Heatmap

```{r}

# Defining genes for visualization
target_genes <- c("KRT1", "KRT2", "KRT9",  
                  "FLG2", "PERP", "DUOX1", 
                  "FOXN1", "CRB3", 
                  "AURKA", "PTPRH", "GPR3", 
                  "FAM81A", "IGHV1-24", "MIR4534", 
                  "FABP7", "SLC35F1", "STEAP1B",
                  "CCDC86", "FGFBP2", "MFAP2", 
                  "VAX2", "SUSD5", "RNU4-1",
                  "RNU4ATAC", "SEC14L4", "GPR50")


# Load precomputed signatures ranked
DE_signatures_fc <- readRDS("DE_signatures_generank.rds")

# Map gene symbols 
map_target_genes <- function(gene_symbols, lookup_df) {
  mapped_genes <- lookup_df %>%
    dplyr::filter(gene_symbol %in% gene_symbols) %>%
    dplyr::select(gene_symbol, ensembl_id) %>%
    distinct()
  
  missing_genes <- setdiff(gene_symbols, mapped_genes$gene_symbol)
  
  if(length(missing_genes) > 0) {
    cat("Warning: Could not find ENSEMBL IDs for:", paste(missing_genes, collapse = ", "), "\n")
  }
  
  cat("Successfully mapped", nrow(mapped_genes), "out of", length(gene_symbols), "genes\n")
  
  return(mapped_genes)
}

target_gene_mapping <- map_target_genes(target_genes, DE_signatures_fc)


# Finding positions and labels of genes in mt
get_target_gene_labels <- function(target_mapping, zscore_matrix) {
  available_genes <- target_mapping[target_mapping$ensembl_id %in% rownames(zscore_matrix), ]
  
  if(nrow(available_genes) == 0) {
    stop("No genes found")
  }
  
  gene_positions <- match(available_genes$ensembl_id, rownames(zscore_matrix))
  gene_labels <- available_genes$gene_symbol
  cat("Found", length(gene_positions), "target genes in the matrix\n")
  cat("Target genes found:", paste(gene_labels, collapse = ", "), "\n")
  
  return(list(
    positions = gene_positions,
    labels = gene_labels,
    ensembl_ids = available_genes$ensembl_id
  ))
}

gene_labels_info <- get_target_gene_labels(target_gene_mapping, data_subset_signatures_z)


gene_text_anno <- rowAnnotation(
  target_genes = anno_mark(
    at = gene_labels_info$positions,
    labels = gene_labels_info$labels,
    labels_gp = gpar(fontsize = 12, fontface = "bold.italic", col = "black"),
    padding = unit(0.5, "cm"),
    link_width = unit(5, "mm"),
    link_gp = gpar(lwd = 1.5, col = "black")
  ),
  width = unit(6, "cm")
)


# Create and save 
set.seed(123)
pdf(height = 20, width = 20, file = "heatmap_cc_signature_across_clusters_040725_new.pdf")

ht_full <- ComplexHeatmap::Heatmap(
  data_subset_signatures_z, 
  #column_split = split,
  column_order = samples_ht,
  cluster_columns = FALSE, 
  top_annotation = colAnn,
  right_annotation = gene_text_anno,  
  column_names_gp = grid::gpar(fontsize = 12),
  column_gap = unit(3, "mm"), 
  show_column_dend = FALSE,
  column_dend_height = unit(5, "cm"),
  show_row_names = FALSE,  
  show_row_dend = FALSE,
  cluster_rows = TRUE,
  clustering_distance_rows = "euclidean",
  clustering_method_rows = "ward.D2",
  name = 'Z-score',
  col = colorRamp2(mybreaks, mycol),
  heatmap_legend_param = list(
    color_bar = 'continuous',
    legend_direction = 'vertical',
    legend_width = unit(8, "cm"),
    legend_height = unit(9, "cm"),
    title_position = 'topcenter',
    title_gp = gpar(fontsize = 15), 
    labels_gp = gpar(fontsize = 12)
  )
)

draw(ht_full, padding = unit(c(2, 2, 2, 2), "cm"),
     heatmap_legend_side = "left")
dev.off()

```



# Boxplots - DE genes across clusters
```{r}
# Define genes of interest for visualization
#genes_of_interest <- c("AURKA", "PTPRH", "GPR3", 
#                       "FAM81A", "IGHV1-24", "MIR4534", 
#                       "FABP7", "SLC35F1", "STEAP1B")

#genes_of_interest <- c("CCDC86", "FGFBP2", "MFAP2", 
#                       "VAX2", "SUSD5", "RNU4-1",
#                       "RNU4ATAC", "SEC14L4", "GPR50")

genes_of_interest <- c("KRT1", "KRT2", "KRT9",
                        "FLG2", "PERP", "DUOX1", 
                       "KPRP", "FOXN1", "CRB3")


# Data preparation
map_genes_to_ensembl <- function(gene_symbols, lookup_df) {
  mapped_genes <- lookup_df %>%
    dplyr::filter(gene_symbol %in% gene_symbols) %>%
    dplyr::select(gene_symbol, ensembl_id) %>%
    distinct()
  
  missing_genes <- setdiff(gene_symbols, mapped_genes$gene_symbol)
  if(length(missing_genes) > 0) {
    cat("Warning: Could not find ENSEMBL IDs for:", paste(missing_genes, collapse = ", "), "\n")
  }
  
  return(mapped_genes)
}

# Map genes
gene_mapping <- map_genes_to_ensembl(genes_of_interest, DE_signatures_fc)
selected_genes_ensembl <- gene_mapping$ensembl_id
names(selected_genes_ensembl) <- gene_mapping$gene_symbol
selected_expr <- mt[selected_genes_ensembl, ]
rownames(selected_expr) <- names(selected_genes_ensembl)

# Prepare for plotting
plot_data <- melt(t(selected_expr))
colnames(plot_data) <- c("Sample", "Gene", "Expression")
plot_data$Sample <- as.character(plot_data$Sample)

# Convert sample cluster classification to df
classification_df <- classification %>% 
  rownames_to_column(var = "Sample")

# Merge with cluster class
plot_data <- merge(plot_data, 
                  classification_df[, c("Sample", "class")], 
                  by = "Sample",
                  all.x = TRUE)
plot_data$Cluster <- as.factor(plot_data$class)


#  Calculate Y-axis limits
y_min <- min(plot_data$Expression, na.rm = TRUE)
y_max <- max(plot_data$Expression, na.rm = TRUE)
y_range <- y_max - y_min
y_buffer <- y_range * 0.60 
shared_ylim <- c(0, 18)

# P-values
get_significance <- function(p) {
  if(p <= 0.0001) return("****")
  if(p <= 0.001) return("***")
  if(p <= 0.01) return("**")
  if(p <= 0.05) return("*")
  return("ns")
}

create_manual_comparison_plot <- function(data, gene_name) {
  gene_data <- data[data$Gene == gene_name, ]
  
  # Statistical tests
  pairwise_tests <- list(
    "1vs2" = wilcox.test(gene_data$Expression[gene_data$Cluster == "1"], 
                        gene_data$Expression[gene_data$Cluster == "2"])$p.value,
    "1vs3" = wilcox.test(gene_data$Expression[gene_data$Cluster == "1"], 
                        gene_data$Expression[gene_data$Cluster == "3"])$p.value,
    "2vs3" = wilcox.test(gene_data$Expression[gene_data$Cluster == "2"], 
                        gene_data$Expression[gene_data$Cluster == "3"])$p.value
  )
  
  sig_symbols <- sapply(pairwise_tests, get_significance)
  
  kw_test <- kruskal.test(Expression ~ Cluster, data = gene_data)
  kw_pvalue <- format.pval(kw_test$p.value, digits = 2)
  
  
#  Bracket positions
gene_data_max <- max(gene_data$Expression, na.rm = TRUE)
plot_height <- shared_ylim[2] - shared_ylim[1] 
bracket_y_base <- gene_data_max + 3.7
max_bracket_y <- bracket_y_base + (2 * plot_height * 0.12) 
if(max_bracket_y > 15) {  
  bracket_y_base <- 17.5 - (2 * plot_height * 0.1) 
}

bracket_spacing <- plot_height * 0.07  


  # Plot
  p <- ggplot(gene_data, aes(x = Cluster, y = Expression, fill = Cluster)) +
    geom_boxplot(alpha = 0.9, 
                 outlier.shape = NA,
                 color = "black",
                 linewidth = 0.6) +
    geom_point(position = position_jitter(width = 0.2, height = 0), 
               alpha = 0.5, 
               size = 2.2, 
               color = "gray40",
               shape = 16) +
    scale_fill_manual(values = c("1" = "#EC99C2", "2" = "#F1955C", "3" = "#A99AEC")) +
    scale_x_discrete(limits = c("1", "2", "3")) +
    scale_y_continuous(limits = shared_ylim, expand = c(0, 0)) +
    labs(title = gene_name) +
    theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      axis.text.x = element_text(size = 12, color = "black"),
      axis.text.y = element_text(size = 12, color = "black"),
      axis.line.x = element_line(color = "black", linewidth = 0.6),
      axis.line.y = element_line(color = "black", linewidth = 0.6),
      axis.ticks.x = element_line(color = "black", linewidth = 0.6),
      axis.ticks.y = element_line(color = "black", linewidth = 0.6),
      axis.ticks.length = unit(0.2, "cm"),
      legend.position = "none",
      panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
      panel.background = element_rect(fill = "white"),
      plot.margin = margin(t = 15, r = 5, b = 5, l = 5)
    ) 
  

  # Cluster 1 vs 2
  bracket1_y <- bracket_y_base
  p <- p +
    annotate("segment", x = 1, xend = 2, 
             y = bracket1_y, yend = bracket1_y, 
             color = "black", linewidth = 0.6) +
    annotate("segment", x = 1, xend = 1, 
             y = bracket1_y - y_range * 0.01, yend = bracket1_y, 
             color = "black", linewidth = 0.6) +
    annotate("segment", x = 2, xend = 2, 
             y = bracket1_y - y_range * 0.01, yend = bracket1_y, 
             color = "black", linewidth = 0.6) 
  
  if(sig_symbols[["1vs2"]] == "ns") {
    p <- p + annotate("text", x = 1.5, y = bracket1_y + y_range * 0.02, 
                     label = sig_symbols[["1vs2"]], size = 3.5, fontface = "plain")
  } else {
    p <- p + annotate("text", x = 1.5, y = bracket1_y + y_range * 0.02, 
                     label = sig_symbols[["1vs2"]], size = 5, fontface = "bold")
  }
  
  
  # Cluster 1 vs 3
  bracket2_y <- bracket1_y + bracket_spacing
  p <- p +
    annotate("segment", x = 1, xend = 3, 
             y = bracket2_y, yend = bracket2_y, 
             color = "black", linewidth = 0.6) +
    annotate("segment", x = 1, xend = 1, 
             y = bracket2_y - y_range * 0.01, yend = bracket2_y, 
             color = "black", linewidth = 0.6) +
    annotate("segment", x = 3, xend = 3, 
             y = bracket2_y - y_range * 0.01, yend = bracket2_y, 
             color = "black", linewidth = 0.6) 
  
  if(sig_symbols[["1vs3"]] == "ns") {
    p <- p + annotate("text", x = 2, y = bracket2_y + y_range * 0.02, 
                     label = sig_symbols[["1vs3"]], size = 3.5, fontface = "plain")
  } else {
    p <- p + annotate("text", x = 2, y = bracket2_y + y_range * 0.02, 
                     label = sig_symbols[["1vs3"]], size = 5, fontface = "bold")
  }
  
  
  # Cluster 2 vs 3
  bracket3_y <- bracket2_y + bracket_spacing
  p <- p +
    annotate("segment", x = 2, xend = 3, 
             y = bracket3_y, yend = bracket3_y, 
             color = "black", linewidth = 0.6) +
    annotate("segment", x = 2, xend = 2, 
             y = bracket3_y - y_range * 0.01, yend = bracket3_y, 
             color = "black", linewidth = 0.6) +
    annotate("segment", x = 3, xend = 3, 
             y = bracket3_y - y_range * 0.01, yend = bracket3_y, 
             color = "black", linewidth = 0.6) 

    if(sig_symbols[["2vs3"]] == "ns") {
    p <- p + annotate("text", x = 2.5, y = bracket3_y + y_range * 0.02, 
                     label = sig_symbols[["2vs3"]], size = 3.5, fontface = "plain")
  } else {
    p <- p + annotate("text", x = 2.5, y = bracket3_y + y_range * 0.02, 
                     label = sig_symbols[["2vs3"]], size = 5, fontface = "bold")
  }
  
  
  p <- p + annotate("text", 
                   x = 1.4, 
                   y = shared_ylim[2] * 0.95,                    
                   label = paste("Kruskal-Wallis, p =", kw_pvalue),
                   size = 3.3, 
                   hjust = 0.5, 
                   fontface = "plain")
  
  return(p)
}

  
# Create plots
manual_plot_list <- list()
for(i in 1:length(genes_of_interest)) {
  gene <- genes_of_interest[i]
  if(gene %in% unique(plot_data$Gene)) {
    manual_plot_list[[i]] <- create_manual_comparison_plot(plot_data, gene)
  } else {
    cat("Warning: Gene", gene, "not found in expression data\n")
    manual_plot_list[[i]] <- NULL
  }
}

manual_plot_list <- manual_plot_list[!sapply(manual_plot_list, is.null)]


# Define layout 
create_publication_layout <- function(plot_list) {
  combined_plot <- grid.arrange(
    grobs = plot_list,
    ncol = 3,
    nrow = 3,
    top = textGrob("Expression of Selected Genes Across Sample Clusters", 
                   gp = gpar(fontsize = 18, fontface = "bold")),
    bottom = textGrob("Cluster", 
                     gp = gpar(fontsize = 14, fontface = "bold")),
    left = textGrob("Gene Expression", 
                   rot = 90, 
                   gp = gpar(fontsize = 14, fontface = "bold"))
  )
  return(combined_plot)
}


# Save 
pdf("gene_expression_across_clusters.pdf", width = 10, height = 12)
final_plot <- create_publication_layout(manual_plot_list)  
dev.off()

final_plot <- create_publication_layout(manual_plot_list) 
cat("\nAnalysis completed successfull!\n")


# Session Info
sessionInfo()
```



**Reference**

-   Gu Z, Schlesner M, Hübschmann D. cola: an R/Bioconductor package for consensus partitioning through a general framework. Nucleic Acids Res. 2021 Feb 22;49(3):e15. doi: 10.1093/nar/gkaa1146. PMID: 33275159; PMCID: PMC7897501.

-   Gu Z, Eils R, Schlesner M. Complex heatmaps reveal patterns and correlations in multidimensional genomic data. Bioinformatics. 2016 Sep 15;32(18):2847-9. doi: 10.1093/bioinformatics/btw313. Epub 2016 May 20. PMID: 27207943.
