---
title: "Acral melanoma MX - consensus clustering"
author: "Estef Vazquez"
output:
  html_document:
    theme: simplex
    toc: yes
    toc_float:
      collapsed: true
---


## Load required libraries

```{r, message=FALSE}
library(cola)
library(ComplexHeatmap)
library(tidyverse)
library(RColorBrewer)
library(colorRamp2)
library(circlize)
```

## Clustering and signature analysis

```{r}
# Load data 
mt <- readRDS("acralMX_expression_data.rds")
metadata <- read_csv("acralMX_metadata_cc.csv")


# Prepare annotation data
metadata <- as.data.frame(metadata)
anno <- metadata %>%
  select(ID, TMB, Whole_Genome_Duplication) %>%
  mutate(Whole_Genome_Duplication = factor(Whole_Genome_Duplication)) 

# Rownames
rownames(anno) <- anno$ID
anno <- anno[,-1]

# Define colors 
colors <- list('Whole_Genome_Duplication' = c('WGD' = 'deeppink4', 'no_WGD' = 'cadetblue1'),
               'TMB' = colorRamp2(c(0, 600), c("white", "red")))

# Check dimensions 
dim(mt)
dim(anno)

# Run consensus clustering or load the precomputed results 
consensus_results = run_all_consensus_partition_methods(mt, anno = muestras_anno, anno_col = colors)

consensus_results <- readRDS("consensus_results.rds")

# Generate report 
cola_report(consensus_results, output_dir= "consensus_clustering_report")

# Extract results for the best performing combination (top-value and partitioning)
SD_PAM_K3 = consensus_results["SD", "pam"]

# Analyze stats and visualize 
get_stats(consensus_results, k = 3)

consensus_heatmap(consensus_results["SD:pam"], k = 3)

collect_classes(consensus_results["SD:pam"])

# PCA
dimension_reduction(SD_PAM_K3, k = 3, method = "PCA")

# Membership heatmap
membership_heatmap(SD_PAM_K3, k = 3)

# Test association with WGD
test_to_known_factors(consensus_results)

# Extract classification
classification <- cbind(get_classes(consensus_results["SD:pam"], k = 3), 
                        get_membership(consensus_results["SD:pam"], k = 3))
write.table(classification, file = "cc_classification.tsv", sep = "\t",  col.names=NA)

# Fix levels for visualization 
split <- factor(paste0("Cluster\n", classification$class), levels=c("Cluster\n1","Cluster\n2","Cluster\n3"))

# Get signatures and perform enrichment analysis 
DE_signatures <- get_signatures(SD_PAM_K3, k = 3)
write.csv(DE_signatures, "signatures.csv")

enrichment_results = functional_enrichment(SD_PAM_K3, k = 3, ontology = c("BP", "MF", "CC")

# Extracting signatures
mt_genes <- rownames(SD_PAM_K3)
sig_index <- DE_signatures$which_row

df <- data.frame(chr=SD_PAM_K3@top_value_list, stringsAsFactors=FALSE)
gene_list <- rownames_to_column(df, var = "ensembl")
gene_signature <- gene_list[sig_index,]
gene_signature <- gene_signature[,-2]
subset_signature <- mt[rownames(mt) %in% gene_signature, ]  
dim(subset_signature)

# Prepare expression data for heatmap
data_subset_signatures_z <- t(scale(t(subset_signature)))

# Or load the precomputed z-score matrix
data_subset_signatures_z <- readRDS("data_subset_signatures_z.rds")

# Prepare metadata
rownames(metadata) <- metadata$ID
metadata <- metadata[,-1]

metadata_ht <- metadata %>% 
  select(BRAF, NRAS, KIT, NF1, Whole_Genome_Duplication, TMB, Ulceration, Breslow, Age, Sex, Stage) %>%
  mutate(across(c(Stage, Sex, Ulceration, Whole_Genome_Duplication, NF1, KIT, NRAS, BRAF), as.factor),
         across(where(is.numeric), as.numeric)) %>% glimpse()


```

## Heatmap visualization

```{r}
ht_opt$ANNOTATION_LEGEND_PADDING = unit(2, "cm")
ht_opt$HEATMAP_LEGEND_PADDING = unit(3, "cm")

# Samples for visualization
samples_ht <- c("PR51978a", "PR51948a", "PR40984a", "PR40966a", "PR41905a", "PR41916a", "PR51951a", "PR40963a", "PR41929d", "PR41045a", "PR41030a", "PR41928d", "PR51928a", "PR41017a", "PR41901a", "PR40985a", "PR41044a", "PR51932a", "PR41011a", "PR41930d", "PR41917a", "PR40989a", "PR41927d", "PR40971a", "PR40962a", "PR40961a", "PR41920a","PR40967a", "PR40996a",  "PR40994a", "PR41020a", "PR40997a", "PR41921a", "PR51979a", "PR51956a", "PR40968a", "PR41002a", "PR41913a", "PR40969a", "PR41025a", "PR40988a", "PR41035a", "PR40970a", "PR40990a", "PR41023f", "PR40986a", "PR41912a")

# Define color functions and breaks
col_fun_tmb = colorRamp2(c(86,310,535), c("white", "#fdf542", "#ff0000"))
col_fun_bres = colorRamp2(c(0,15,30), c("white", "#fdf542", "#ff0000"))
col_fun_age = colorRamp2(c(32,60,88), c("white", "#fdf542", "#ff0000"))
mycol <- colorRampPalette(c('green', 'black', 'purple'))(100)
mybreaks <- seq(-2, 2, length.out = 100) 

colors <- list('BRAF' = c('Missense' = 'red', 'Non_mutated' = 'blue4'),
               'NRAS' = c('Missense' = 'olivedrab', 'Non_mutated' = 'orange'), 
               'KIT' = c('Missense' = '#3F007D', 'no-SNV' = 'yellow', 'Non_mutated' = '#CC0066'),
               'NF1' = c("Nonsense" = "#CB181D", 'no-SNV' = '#74a9cf', 'Non_mutated' = 'pink'),
               'Whole_Genome_Duplication' = c('WGD' = 'deeppink4', 'no_WGD' = 'cadetblue1'),
               'TMB' = col_fun_tmb,
               'Ulceration' = c('yes' = '#00441B', 'no' = '#74a9cf'),
               'Breslow' = col_fun_bres,
               'Age' = col_fun_age,
               'Sex' = c('female' = 'blue', 'male' = '#241178'),
               'Stage' = c('0' = '#FFFFCC', '1' = '#FFFF66', '2' = '#FF9900', '3' = 'red', '4' = "gray"))

# Annotations
colAnn <- HeatmapAnnotation(
  df = metadata_ht,
  which = 'col', 
  na_col = "black", 
  col = colors,
  simple_anno_size = unit(1.4, "cm"),
  annotation_width = unit(c(1, 11), 'cm'),
  gap = unit(1.5, 'mm'),
  annotation_height = unit(c(1, 17), 'cm'),
  annotation_name_gp = gpar(fontsize = 21),
  annotation_name_align = TRUE,
  annotation_name_side = 'left',
  annotation_legend_param = list(
      BRAF = list(
        #nrow = 2, 
        title = 'BRAF',
        title_gp = gpar(fontsize = 32, fontface = 'bold'),
        labels_gp = gpar(fontsize = 32)),
      NRAS = list(
        #nrow = 2,
        title = 'NRAS',
        title_gp = gpar(fontsize = 32, fontface = 'bold'),
        labels_gp = gpar(fontsize = 32)),
      KIT = list(
        #nrow = 3,
        title = 'KIT',
        title_gp = gpar(fontsize = 32, fontface = 'bold'),
        labels_gp = gpar(fontsize = 32)),
      NF1 = list(
        #nrow = 3,
        title = 'NF1',
        title_gp = gpar(fontsize = 32, fontface = 'bold'),
        labels_gp = gpar(fontsize = 32)),
      Whole_Genome_Duplication = list(
        #nrow = 2,
        title = 'WGD',
        title_gp = gpar(fontsize = 32, fontface = 'bold'),
        labels_gp = gpar(fontsize = 32)),
      TMB = list(
        #nrow = 2,
        title = 'TMB',
        title_gp = gpar(fontsize = 32, fontface = 'bold'),
        labels_gp = gpar(fontsize = 32)),
      Age = list(
        #nrow = 2,
        title = 'Age',
        title_gp = gpar(fontsize = 32, fontface = 'bold'),
        labels_gp = gpar(fontsize = 32)),
      Sex = list(
        #nrow = 2,
        title = 'Sex',
        title_gp = gpar(fontsize = 32, fontface = 'bold'),
        labels_gp = gpar(fontsize = 32)),
      Breslow = list(
        #nrow = 2,
        title = 'Breslow',
        title_gp = gpar(fontsize = 32, fontface = 'bold'),
        labels_gp = gpar(fontsize = 32)),
      Ulceration = list(
        #nrow = 3,
        title = 'Ulceration',
        at = c("yes", "no"),
        labels = c("Yes", "No"),
        title_gp = gpar(fontsize = 32, fontface = 'bold'),
        labels_gp = gpar(fontsize = 32)),
      Stage = list(
        #nrow = 4,
        title = 'Stage',
        title_gp = gpar(fontsize = 32, fontface = 'bold'),
        labels_gp = gpar(fontsize = 32))))
      
# Gene labels
genelabels <- rowAnnotation(
  Genes = anno_mark(
    at = seq(1, nrow(data_subset_signatures_z), 100),
    labels = rownames(data_subset_signatures_z)[seq(1, nrow(data_subset_signatures_z), 100)],
    labels_gp = gpar(fontsize = 32),
    padding = 1),
    width = unit(14.0, 'cm') + 
    max_text_width(
  rownames(data_subset_signatures_z)[seq(1, nrow(data_subset_signatures_z), 100)],
  gp = gpar(fontsize= 32, fontface = 'bold')))


# Save heatmap 
set.seed(123)
png(height=3100, width=3100, file="heatmap.png", type = "cairo")

ht <- ComplexHeatmap::Heatmap(data_subset_signatures_z,
                                      column_split=split, 
                                      raster_by_magick = FALSE,
                                      column_order = samples_ht, 
                                      top_annotation = colAnn,
                                      column_names_gp = grid::gpar(fontsize = 20, fontface = 'bold'),
                                      column_gap = unit(5, "mm"), 
                                      show_column_dend = TRUE,
                                      column_dend_height = unit(17,'cm'),
                                      show_row_names = FALSE,
                                      show_row_dend = FALSE,
                                      cluster_rows = TRUE,
                                      clustering_distance_rows = "euclidean", 
                                      clustering_method_rows = "ward.D2", 
                                      name = 'Z-score',
                                      col = colorRamp2(mybreaks, mycol),
                                      heatmap_legend_param = list(
                                        color_bar = 'continuous',
                                        legend_direction = 'vertical',
                                        legend_width = unit(30, 'cm'),
                                        legend_height = unit(25, 'cm'),
                                        at = c(-3, 0, 3),
                                        title_gp=gpar(fontsize = 50),
                                        labels_gp=gpar(fontsize = 50)))


ht = draw(ht)
dev.off()
```


**Reference**

-   Gu Z, Schlesner M, Hübschmann D. cola: an R/Bioconductor package for consensus partitioning through a general framework. Nucleic Acids Res. 2021 Feb 22;49(3):e15. doi: 10.1093/nar/gkaa1146. PMID: 33275159; PMCID: PMC7897501.

-   Gu Z, Eils R, Schlesner M. Complex heatmaps reveal patterns and correlations in multidimensional genomic data. Bioinformatics. 2016 Sep 15;32(18):2847-9. doi: 10.1093/bioinformatics/btw313. Epub 2016 May 20. PMID: 27207943.
