---
title: "Acral melanoma MX - consensus clustering analysis, enrichment and visualization of gene signatures across clusters"
author: "Estef Vazquez"
output:
  html_document:
    theme: simplex
    toc: yes
    toc_float:
      collapsed: true
---

Overview
This document provides code for running consensus clustering to identify molecular subtypes (two-level clustering approach, sample classification followed by genes), differential expression, enrichment analysis and visualization of gene expression signatures in 44 acral melanoma samples. The analysis also generates a heatmap showing expression patterns and boxplot comparisons of differentially expressed genes across clusters.


## Load required libraries

```{r, message=FALSE}

library(here)
library(cola)
library(ComplexHeatmap)
library(clusterProfiler)
library(tidyverse)
library(ggpubr)
library(RColorBrewer)
library(colorRamp2)
library(circlize)
library(cowplot)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(grid)
library(gridExtra)
library(reshape2)
```


## Clustering and signature analysis

```{r}
# Load data 
mt <- readRDS("acralMX_expression_data_update44s.rds")
metadata <- read_csv("acralMX_metadata_cc_update.csv")


# Prepare annotation data
metadata <- as.data.frame(metadata)
cc_anno <- metadata %>%
  select(TMB, WGD) %>%
  mutate(WGD = factor(WGD)) 

# Rownames
rownames(cc_anno) <- cc_anno$sample_id
cc_anno <- cc_anno[,-1]

# Define colors 
colors <- list('WGD' = c('WGD' = 'deeppink4', 'no_WGD' = 'cadetblue1'),
               'TMB' = colorRamp2(c(0, 600), c("white", "red")))

# Check dimensions 
dim(mt)
dim(anno)

# Run consensus clustering 
consensus_results = run_all_consensus_partition_methods(mt, anno = cc_anno, anno_col = colors)

# Generate report 
cola_report(consensus_results, output_dir= "consensus_clustering_report")

# Extract results for the best performing combination (top-value and partitioning)
SD_PAM_K3 = consensus_results["SD", "pam"]

# Analyze stats and visualize 
get_stats(consensus_results, k = 3)

consensus_heatmap(consensus_results["SD:pam"], k = 3)
collect_classes(consensus_results["SD:pam"])

# PCA
dimension_reduction(SD_PAM_K3, k = 3, method = "PCA")

# Membership heatmap
membership_heatmap(SD_PAM_K3, k = 3)

# Test association with WGD
test_to_known_factors(consensus_results)

# Extract classification
classification <- cbind(get_classes(consensus_results["SD:pam"], k = 3), 
                        get_membership(consensus_results["SD:pam"], k = 3))
write.table(classification, file = "cc_classification.tsv", sep = "\t",  col.names=NA)

# Fix levels for visualization 
split <- factor(paste0("Cluster\n", classification$class), levels=c("Cluster\n1","Cluster\n2","Cluster\n3"))

# Get signatures and perform enrichment analysis 
DE_signatures <- get_signatures(SD_PAM_K3, k = 3)
write.csv(DE_signatures, "signatures.csv")

enrichment_results = functional_enrichment(SD_PAM_K3, k = 3, ontology = c("BP", "MF", "CC")

# Extracting signatures
mt_genes <- rownames(SD_PAM_K3)
sig_index <- DE_signatures$which_row

df <- data.frame(chr=SD_PAM_K3@top_value_list, stringsAsFactors=FALSE)
gene_list <- rownames_to_column(df, var = "ensembl")
gene_signature <- gene_list[sig_index,]
gene_signature <- gene_signature[,-2]
subset_signature <- mt[rownames(mt) %in% gene_signature, ]  
dim(subset_signature)

# Prepare expression data for heatmap
data_subset_signatures_z <- t(scale(t(subset_signature)))

# Or load the precomputed z-score matrix
data_subset_signatures_z <- readRDS("DE_signatures_zscores.rds")
```


## Heatmap visualization

```{r}

# Verify samples in mt and metadata are present
all_expr_samples_in_metadata <- all(colnames(data_subset_signatures_z) %in% rownames(metadata))
print(paste("All expression samples in metadata:", all_expr_samples_in_metadata))
all_metadata_samples_in_expr <- all(rownames(metadata) %in% colnames(data_subset_signatures_z))
print(paste("All metadata samples in expression data:", all_metadata_samples_in_expr))

# Verify metadata order 
metadata_ordered <- metadata[colnames(data_subset_signatures_z), ]
identical(rownames(metadata_ordered), colnames(data_subset_signatures_z))

# Samples 
samples_ht <- c("PR51978a", "PR51948a", "PR40984a", "PR40966a", "PR41905a", "PR41916a", "PR51951a", "PR40963a", "PR41045a", "PR40996a", "PR51979a", 
                 "PR41023f", "PR40986a", "PR41912a", "PR41928d", "PR51928a", "PR41017a", "PR41901a", "PR40985a", "PR41044a", "PR51932a", "PR41011a",
                 "PR41930d", "PR40989a", "PR41927d", "PR40971a", "PR40962a", "PR40961a", "PR41920a", "PR40967a", "PR41030a", "PR40994a", "PR41020a",     
                 "PR40997a", "PR41921a", "PR40968a", "PR41002a", "PR41913a", "PR40969a", "PR41025a", "PR40988a", "PR41035a", "PR40970a", "PR40990a")

# Verify class order
classification_ordered <- classification[samples_ht, ]
table(classification_ordered$class)
nrow(classification_ordered)
length(samples_ht) 

# Create split
split <- factor(paste0("Cluster", classification_ordered$class), 
                levels=c("Cluster1", "Cluster2", "Cluster3"))
names(split) <- samples_ht


# Verify data is aligned
data_subset_signatures_z <- data_subset_signatures_z[, samples_ht]
identical(names(split), samples_ht)  
identical(names(split), colnames(data_subset_signatures_z))  
metadata_ordered <- metadata_ordered[samples_ht, ]
metadata_ordered <- metadata[samples_ht, ]

identical(rownames(metadata_ordered), samples_ht)  
identical(rownames(metadata_ordered), colnames(data_subset_signatures_z)) 


# Annotations for heatmap 

# TMB
tmb_breaks <- c(min(metadata_ordered$TMB, na.rm=TRUE), 
                median(metadata_ordered$TMB, na.rm=TRUE), 
                max(metadata_ordered$TMB, na.rm=TRUE))

col_fun_tmb = colorRamp2(tmb_breaks, c("white", "#fdf542", "#ff0000"))


# Age
age_breaks <- c(min(metadata_ordered$age, na.rm=TRUE),
                median(metadata_ordered$age, na.rm=TRUE),
                max(metadata_ordered$age, na.rm=TRUE))

col_fun_age = colorRamp2(age_breaks, c("#FFEDAB", "#FFA500", "#CD6600"))


# Tumor purity
col_fun_purity = colorRamp2(c(0, 0.5, 1.0), c("white", "#fdf542", "#ff0000"))


# Colors HT
mycol <- colorRampPalette(c('green', 'black', 'purple'))(100)
mybreaks <- seq(-2, 2, length.out = 100) 

colors <- list(
  'mutation_status' = c('BRAF' = '#F48FB1', 'NF1' = '#81C784', 
                        'NRAS' = '#FFAB91', 'KIT' = '#80DEEA',
                        'Other' = '#BDBDBD', 'multihit' = '#FFF176'), 
  'stage' = c('0' = '#E0E0E0', '1' = '#BBDEFB', '2' = '#90CAF9',
              '3' = '#5C6BC0', '4' = '#4527A0'),
  'sex' = c('F' = '#4DAF4A', 'M' = '#FC4E2A'),
  'age' = col_fun_age,
  'ulceration' = c('1' = '#CB181D', '0' = '#2196F3'),
  'WGD' = c('1' = 'deeppink4', '0' = '#4FC3F7'),
  'TMB' = col_fun_tmb,
  'tumor_purity' = col_fun_purity
)


colAnn <- HeatmapAnnotation(
  df = metadata_ordered,
  which = 'col', 
  na_col = "black", 
  col = colors,
  simple_anno_size = unit(0.5, "cm"),
  annotation_width = unit(c(1, 8), 'cm'),  # Adjusted for 8 columns
  gap = unit(0.5, 'mm'),
  annotation_height = unit(0.5, 'cm'),  # Slightly reduced for fewer columns
  annotation_name_gp = gpar(fontsize = 15),
  annotation_name_align = TRUE,
  annotation_name_side = 'left',
  annotation_legend_param = list(
      mutation_status = list(
        title = 'Mutation Status',
        title_gp = gpar(fontsize = 22, fontface = 'bold'),
        labels_gp = gpar(fontsize = 20)),
      stage = list(
        title = 'Stage',
        title_gp = gpar(fontsize = 22, fontface = 'bold'),
        labels_gp = gpar(fontsize = 20)),
      sex = list(
        title = 'Sex',
        title_gp = gpar(fontsize = 22, fontface = 'bold'),
        labels_gp = gpar(fontsize = 20)),
      age = list(
        title = 'Age',
        title_gp = gpar(fontsize = 22, fontface = 'bold'),
        labels_gp = gpar(fontsize = 20)),
      ulceration = list(
        title = 'Ulceration',
        at = c("1", "0"),  # Adapt based on your actual values
        labels = c("Yes", "No"),
        title_gp = gpar(fontsize = 22, fontface = 'bold'),
        labels_gp = gpar(fontsize = 20)),
      WGD = list(
        title = 'WGD',
        at = c("1", "0"),  # Adapt based on your actual values
        labels = c("Yes", "No"),
        title_gp = gpar(fontsize = 22, fontface = 'bold'),
        labels_gp = gpar(fontsize = 20)),
      TMB = list(
        title = 'TMB',
        title_gp = gpar(fontsize = 22, fontface = 'bold'),
        labels_gp = gpar(fontsize = 20)),
      tumor_purity = list(
        title = 'Tumor Purity',
        title_gp = gpar(fontsize = 22, fontface = 'bold'),
        labels_gp = gpar(fontsize = 20))
  )
)
```


# Clustering DE signatures - Heatmap

```{r}

# Defining genes for visualization
target_genes <- c("KRT1", "KRT2", "KRT9",  
                  "FLG2", "PERP", "DUOX1", 
                  "FOXN1", "CRB3", 
                  "AURKA", "PTPRH", "GPR3", 
                  "FAM81A", "IGHV1-24", "MIR4534", 
                  "FABP7", "SLC35F1", "STEAP1B",
                  "CCDC86", "FGFBP2", "MFAP2", 
                  "VAX2", "SUSD5", "RNU4-1",
                  "RNU4ATAC", "SEC14L4", "GPR50")

# Load signatures ranked
DE_signatures_fc <- readRDS("DE_signatures_generank.rds")


# Map gene symbols 
map_target_genes <- function(gene_symbols, lookup_df) {
  mapped_genes <- lookup_df %>%
    dplyr::filter(gene_symbol %in% gene_symbols) %>%
    dplyr::select(gene_symbol, ensembl_id) %>%
    distinct()
  
  missing_genes <- setdiff(gene_symbols, mapped_genes$gene_symbol)
  
  if(length(missing_genes) > 0) {
    cat("Warning: Could not find ENSEMBL IDs for:", paste(missing_genes, collapse = ", "), "\n")
  }
  
  cat("Successfully mapped", nrow(mapped_genes), "out of", length(gene_symbols), "genes\n")
  
  return(mapped_genes)
}

target_gene_mapping <- map_target_genes(target_genes, DE_signatures_fc)



# Finding positions and labels of genes in mt
get_target_gene_labels <- function(target_mapping, zscore_matrix) {
  available_genes <- target_mapping[target_mapping$ensembl_id %in% rownames(zscore_matrix), ]
  
  if(nrow(available_genes) == 0) {
    stop("No genes found")
  }
  
  gene_positions <- match(available_genes$ensembl_id, rownames(zscore_matrix))
  gene_labels <- available_genes$gene_symbol
  cat("Found", length(gene_positions), "target genes in the matrix\n")
  cat("Target genes found:", paste(gene_labels, collapse = ", "), "\n")
  
  return(list(
    positions = gene_positions,
    labels = gene_labels,
    ensembl_ids = available_genes$ensembl_id
  ))
}

gene_labels_info <- get_target_gene_labels(target_gene_mapping, data_subset_signatures_z)


# Gene annotation
gene_text_anno <- rowAnnotation(
  target_genes = anno_mark(
    at = gene_labels_info$positions,
    labels = gene_labels_info$labels,
    labels_gp = gpar(fontsize = 12, fontface = "bold.italic", col = "black"),
    padding = unit(0.5, "cm"),
    link_width = unit(5, "mm"),
    link_gp = gpar(lwd = 1.5, col = "black")
  ),
  width = unit(6, "cm")
)


# Create and save heatmap
set.seed(123)
pdf(height = 20, width = 20, file = "heatmap_cc_signature_across_clusters_040725_new.pdf")

ht_full <- ComplexHeatmap::Heatmap(
  data_subset_signatures_z, 
  #column_split = split,
  column_order = samples_ht,
  cluster_columns = FALSE, 
  top_annotation = colAnn,
  right_annotation = gene_text_anno,  
  column_names_gp = grid::gpar(fontsize = 12),
  column_gap = unit(3, "mm"), 
  show_column_dend = FALSE,
  column_dend_height = unit(5, "cm"),
  show_row_names = FALSE,  
  show_row_dend = FALSE,
  cluster_rows = TRUE,
  clustering_distance_rows = "euclidean",
  clustering_method_rows = "ward.D2",
  name = 'Z-score',
  col = colorRamp2(mybreaks, mycol),
  heatmap_legend_param = list(
    color_bar = 'continuous',
    legend_direction = 'vertical',
    legend_width = unit(8, "cm"),
    legend_height = unit(9, "cm"),
    title_position = 'topcenter',
    title_gp = gpar(fontsize = 15), 
    labels_gp = gpar(fontsize = 12)
  )
)

draw(ht_full, padding = unit(c(2, 2, 2, 2), "cm"),
     heatmap_legend_side = "left")
dev.off()

```



# Boxplots - DE genes across clusters
```{r}
# Define genes of interest for visualization
#genes_of_interest <- c("AURKA", "PTPRH", "GPR3", 
#                       "FAM81A", "IGHV1-24", "MIR4534", 
#                       "FABP7", "SLC35F1", "STEAP1B")

#genes_of_interest <- c("CCDC86", "FGFBP2", "MFAP2", 
#                       "VAX2", "SUSD5", "RNU4-1",
#                       "RNU4ATAC", "SEC14L4", "GPR50")

genes_of_interest <- c("KRT1", "KRT2", "KRT9",
                        "FLG2", "PERP", "DUOX1", 
                       "KPRP", "FOXN1", "CRB3")



# Data preparation
map_genes_to_ensembl <- function(gene_symbols, lookup_df) {
  mapped_genes <- lookup_df %>%
    dplyr::filter(gene_symbol %in% gene_symbols) %>%
    dplyr::select(gene_symbol, ensembl_id) %>%
    distinct()
  
  missing_genes <- setdiff(gene_symbols, mapped_genes$gene_symbol)
  if(length(missing_genes) > 0) {
    cat("Warning: Could not find ENSEMBL IDs for:", paste(missing_genes, collapse = ", "), "\n")
  }
  
  return(mapped_genes)
}


# Map genes
gene_mapping <- map_genes_to_ensembl(genes_of_interest, DE_signatures_fc)
selected_genes_ensembl <- gene_mapping$ensembl_id
names(selected_genes_ensembl) <- gene_mapping$gene_symbol
selected_expr <- mt[selected_genes_ensembl, ]
rownames(selected_expr) <- names(selected_genes_ensembl)


# Prepare for plotting
plot_data <- melt(t(selected_expr))
colnames(plot_data) <- c("Sample", "Gene", "Expression")
plot_data$Sample <- as.character(plot_data$Sample)
#classification <- classification %>% rownames_to_column(var = "Sample") %>% glimpse()


# Merge with cluster class
plot_data <- merge(plot_data, 
                  classification[, c("Sample", "class")], 
                  by = "Sample",
                  all.x = TRUE)

plot_data$Cluster <- as.factor(plot_data$class)


#  Calculate Y-axis limits
y_min <- min(plot_data$Expression, na.rm = TRUE)
y_max <- max(plot_data$Expression, na.rm = TRUE)
y_range <- y_max - y_min
y_buffer <- y_range * 0.60 
shared_ylim <- c(0, 18)

# P-values
get_significance <- function(p) {
  if(p <= 0.0001) return("****")
  if(p <= 0.001) return("***")
  if(p <= 0.01) return("**")
  if(p <= 0.05) return("*")
  return("ns")
}


create_manual_comparison_plot <- function(data, gene_name) {
  gene_data <- data[data$Gene == gene_name, ]
  
  # Statistical tests
  pairwise_tests <- list(
    "1vs2" = wilcox.test(gene_data$Expression[gene_data$Cluster == "1"], 
                        gene_data$Expression[gene_data$Cluster == "2"])$p.value,
    "1vs3" = wilcox.test(gene_data$Expression[gene_data$Cluster == "1"], 
                        gene_data$Expression[gene_data$Cluster == "3"])$p.value,
    "2vs3" = wilcox.test(gene_data$Expression[gene_data$Cluster == "2"], 
                        gene_data$Expression[gene_data$Cluster == "3"])$p.value
  )
  
  sig_symbols <- sapply(pairwise_tests, get_significance)
  
  kw_test <- kruskal.test(Expression ~ Cluster, data = gene_data)
  kw_pvalue <- format.pval(kw_test$p.value, digits = 2)
  
  
#  Bracket positions
gene_data_max <- max(gene_data$Expression, na.rm = TRUE)
plot_height <- shared_ylim[2] - shared_ylim[1] 
bracket_y_base <- gene_data_max + 3.7
max_bracket_y <- bracket_y_base + (2 * plot_height * 0.12) 
if(max_bracket_y > 15) {  
  bracket_y_base <- 17.5 - (2 * plot_height * 0.1) 
}

bracket_spacing <- plot_height * 0.07  



  # Plot
  p <- ggplot(gene_data, aes(x = Cluster, y = Expression, fill = Cluster)) +
    geom_boxplot(alpha = 0.9, 
                 outlier.shape = NA,
                 color = "black",
                 linewidth = 0.6) +
    geom_point(position = position_jitter(width = 0.2, height = 0), 
               alpha = 0.5, 
               size = 2.2, 
               color = "gray40",
               shape = 16) +
    scale_fill_manual(values = c("1" = "#EC99C2", "2" = "#F1955C", "3" = "#A99AEC")) +
    scale_x_discrete(limits = c("1", "2", "3")) +
    scale_y_continuous(limits = shared_ylim, expand = c(0, 0)) +
    labs(title = gene_name) +
    theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      axis.text.x = element_text(size = 12, color = "black"),
      axis.text.y = element_text(size = 12, color = "black"),
      axis.line.x = element_line(color = "black", linewidth = 0.6),
      axis.line.y = element_line(color = "black", linewidth = 0.6),
      axis.ticks.x = element_line(color = "black", linewidth = 0.6),
      axis.ticks.y = element_line(color = "black", linewidth = 0.6),
      axis.ticks.length = unit(0.2, "cm"),
      legend.position = "none",
      panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
      panel.background = element_rect(fill = "white"),
      plot.margin = margin(t = 15, r = 5, b = 5, l = 5)
    ) 
  

  # Cluster 1 vs 2
  bracket1_y <- bracket_y_base
  p <- p +
    annotate("segment", x = 1, xend = 2, 
             y = bracket1_y, yend = bracket1_y, 
             color = "black", linewidth = 0.6) +
    annotate("segment", x = 1, xend = 1, 
             y = bracket1_y - y_range * 0.01, yend = bracket1_y, 
             color = "black", linewidth = 0.6) +
    annotate("segment", x = 2, xend = 2, 
             y = bracket1_y - y_range * 0.01, yend = bracket1_y, 
             color = "black", linewidth = 0.6) 
  
  if(sig_symbols[["1vs2"]] == "ns") {
    p <- p + annotate("text", x = 1.5, y = bracket1_y + y_range * 0.02, 
                     label = sig_symbols[["1vs2"]], size = 3.5, fontface = "plain")
  } else {
    p <- p + annotate("text", x = 1.5, y = bracket1_y + y_range * 0.02, 
                     label = sig_symbols[["1vs2"]], size = 5, fontface = "bold")
  }
  
  
  # Cluster 1 vs 3
  bracket2_y <- bracket1_y + bracket_spacing
  p <- p +
    annotate("segment", x = 1, xend = 3, 
             y = bracket2_y, yend = bracket2_y, 
             color = "black", linewidth = 0.6) +
    annotate("segment", x = 1, xend = 1, 
             y = bracket2_y - y_range * 0.01, yend = bracket2_y, 
             color = "black", linewidth = 0.6) +
    annotate("segment", x = 3, xend = 3, 
             y = bracket2_y - y_range * 0.01, yend = bracket2_y, 
             color = "black", linewidth = 0.6) 
  
  if(sig_symbols[["1vs3"]] == "ns") {
    p <- p + annotate("text", x = 2, y = bracket2_y + y_range * 0.02, 
                     label = sig_symbols[["1vs3"]], size = 3.5, fontface = "plain")
  } else {
    p <- p + annotate("text", x = 2, y = bracket2_y + y_range * 0.02, 
                     label = sig_symbols[["1vs3"]], size = 5, fontface = "bold")
  }
  
  
  # Cluster 2 vs 3
  bracket3_y <- bracket2_y + bracket_spacing
  p <- p +
    annotate("segment", x = 2, xend = 3, 
             y = bracket3_y, yend = bracket3_y, 
             color = "black", linewidth = 0.6) +
    annotate("segment", x = 2, xend = 2, 
             y = bracket3_y - y_range * 0.01, yend = bracket3_y, 
             color = "black", linewidth = 0.6) +
    annotate("segment", x = 3, xend = 3, 
             y = bracket3_y - y_range * 0.01, yend = bracket3_y, 
             color = "black", linewidth = 0.6) 

    if(sig_symbols[["2vs3"]] == "ns") {
    p <- p + annotate("text", x = 2.5, y = bracket3_y + y_range * 0.02, 
                     label = sig_symbols[["2vs3"]], size = 3.5, fontface = "plain")
  } else {
    p <- p + annotate("text", x = 2.5, y = bracket3_y + y_range * 0.02, 
                     label = sig_symbols[["2vs3"]], size = 5, fontface = "bold")
  }
  
  
  p <- p + annotate("text", 
                   x = 1.4, 
                   y = shared_ylim[2] * 0.95,                    
                   label = paste("Kruskal-Wallis, p =", kw_pvalue),
                   size = 3.3, 
                   hjust = 0.5, 
                   fontface = "plain")
  
  return(p)
}

  
# Create plots
manual_plot_list <- list()
for(i in 1:length(genes_of_interest)) {
  gene <- genes_of_interest[i]
  if(gene %in% unique(plot_data$Gene)) {
    manual_plot_list[[i]] <- create_manual_comparison_plot(plot_data, gene)
  } else {
    cat("Warning: Gene", gene, "not found in expression data\n")
    manual_plot_list[[i]] <- NULL
  }
}

manual_plot_list <- manual_plot_list[!sapply(manual_plot_list, is.null)]


# Define layout 
create_publication_layout <- function(plot_list) {
  combined_plot <- grid.arrange(
    grobs = plot_list,
    ncol = 3,
    nrow = 3,
    top = textGrob("Expression of Selected Genes Across Sample Clusters", 
                   gp = gpar(fontsize = 18, fontface = "bold")),
    bottom = textGrob("Cluster", 
                     gp = gpar(fontsize = 14, fontface = "bold")),
    left = textGrob("Gene Expression", 
                   rot = 90, 
                   gp = gpar(fontsize = 14, fontface = "bold"))
  )
  return(combined_plot)
}


# Save 
pdf("gene_expression_across_clusters.pdf", width = 10, height = 12)
final_plot <- create_publication_layout(manual_plot_list)  
dev.off()

final_plot <- create_publication_layout(manual_plot_list) 
```



**Reference**

-   Gu Z, Schlesner M, Hübschmann D. cola: an R/Bioconductor package for consensus partitioning through a general framework. Nucleic Acids Res. 2021 Feb 22;49(3):e15. doi: 10.1093/nar/gkaa1146. PMID: 33275159; PMCID: PMC7897501.

-   Gu Z, Eils R, Schlesner M. Complex heatmaps reveal patterns and correlations in multidimensional genomic data. Bioinformatics. 2016 Sep 15;32(18):2847-9. doi: 10.1093/bioinformatics/btw313. Epub 2016 May 20. PMID: 27207943.
